---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# cassidyr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/JDenn0514/cassidyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JDenn0514/cassidyr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**cassidyr** is an R client for the [CassidyAI](https://www.cassidyai.com/) API, providing a clean and intuitive interface to integrate AI-powered assistants into your R workflows.

## Features

- âœ¨ **Unified Console Chat**: Just call `cassidy_chat()` - conversations are automatically managed, persisted, and resumable
- ðŸ¤– **Simple API Client**: Create threads, send messages, and manage conversations
- ðŸ“Š **Context-Aware**: Automatically gather project and data context for better AI responses
- ðŸ’¬ **Interactive Chat**: Launch a Shiny-based chat interface with `cassidy_app()`
- ðŸ’¾ **Conversation Persistence**: Auto-save and restore chat history
- ðŸ“‹ **Copy Code Button**: One-click copy for code blocks in chat responses
- ðŸŽ¯ **Token Management**: Automatic tracking, warnings, and compaction to prevent limit errors
- ðŸ’­ **Memory System**: Persistent knowledge storage for unlimited conversation length
- ðŸ”§ **IDE Integration** (coming soon): RStudio/Positron addins for AI-assisted coding

## Installation

You can install the development version of cassidyr from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("JDenn0514/cassidyr")
```

## Setup

Before using cassidyr, you'll need:
1. **CassidyAI Account**: Sign up at [cassidyai.com](https://www.cassidyai.com/)
2. **API Key**: Generate one from Organization Settings â†’ API Keys
3. **Assistant ID**: Find this in your assistant's External Deployments settings

Store your credentials securely in your `.Renviron` file:

```{r, eval = FALSE}
# Open your .Renviron file
usethis::edit_r_environ()

# Add these lines (with your actual credentials):
# CASSIDY_API_KEY = "your_api_key_here"
# CASSIDY_ASSISTANT_ID = "your_assistant_id_here"

# Save, close, and restart R
```
## Quick Start

### Interactive Chat App

The easiest way to use cassidyr is through the interactive Shiny chat interface:

```{r, eval = FALSE}
library(cassidyr)

# Launch the chat app
cassidy_app()
```

Features include:
- Context sidebar for managing project/data/file context
- Conversation history with save/load
- One-click code copying
- Mobile-responsive design

### Console Chat (New!)

The unified console interface makes interactive R sessions effortless - just chat!

```{r, eval = FALSE}
library(cassidyr)

# Just start chatting - conversations are automatically managed
cassidy_chat("How do I create a bar plot in ggplot2?")
cassidy_chat("Show me with the mtcars dataset")  # Automatically continues

# Start a new conversation
cassidy_chat("New topic: What is the tidyverse?", conversation = "new")

# Check your conversations
cassidy_conversations()

# Switch between conversations
convs <- cassidy_conversations()
cassidy_chat("Let's continue", conversation = convs$id[1])

# Reset to start fresh
cassidy_reset()
```

For programmatic use (scripts, Quarto documents), session objects provide more control:

```{r, eval = FALSE}
# Session-based approach (recommended for scripts)
session <- cassidy_session()
chat(session, "What is the tidyverse?")
chat(session, "Which package should I use for data cleaning?")
```

### Low-Level API Access

For more control over thread management:

```{r, eval = FALSE}
# Create a thread
thread_id <- cassidy_create_thread()

# Send messages
response <- cassidy_send_message(thread_id, "What is the tidyverse?")
print(response)

# Retrieve conversation history
thread <- cassidy_get_thread(thread_id)

# List all threads
threads <- cassidy_list_threads()
```

### Context Gathering

Provide rich context to improve AI responses. cassidyr's memory system (inspired by [Claude Code](https://code.claude.com/docs/en/memory)) searches recursively up your directory tree, combining user-level preferences with project-specific settings:

```{r, eval = FALSE}
# Create a CASSIDY.md file for your project
use_cassidy_md()  # Creates CASSIDY.md with project instructions

# Gather comprehensive project context (includes CASSIDY.md files)
ctx <- cassidy_context_project(level = "standard")

# Describe a data frame for the AI
cassidy_describe_df(mtcars)

# Files are automatically loaded when you start cassidy_app()
```

Memory files can be placed at multiple levels:
- **User-level**: `~/.cassidy/CASSIDY.md` (applies to all projects)
- **Project-level**: `CASSIDY.md` or `.cassidy/CASSIDY.md` (shared with team)
- **Local**: `CASSIDY.local.md` (personal, auto-gitignored)
- **Modular**: `.cassidy/rules/*.md` (organized by topic)

## Managing Long Conversations

The CassidyAI API has a 200,000 token limit. cassidyr provides comprehensive tools to manage long conversations:

### Automatic Token Tracking

Token usage is tracked automatically across all interfaces:

```{r, eval = FALSE}
# Console chat shows token warnings
cassidy_chat("Let's have a long conversation...")
#> Tokens: 12,450 / 200,000 (6%)

# Session objects track tokens
session <- cassidy_session()
session <- chat(session, "Hello!")
print(session)
#> Cassidy Session
#> Thread ID: abc123
#> Messages: 2
#> Tokens: 1,234 / 200,000 (0.6%)

# Get detailed statistics
stats <- cassidy_session_stats(session)
```

### Automatic Compaction

`cassidy_session()` automatically compacts conversations before hitting limits:

```{r, eval = FALSE}
# Auto-compaction enabled by default
session <- cassidy_session()

# Conversation automatically compacts at 85% token usage
session <- chat(session, "Long message...")
#> Warning: Approaching token limit, auto-compacting...
#> Compaction complete. Continuing with your message...

# Customize threshold or disable
session <- cassidy_session(
  compact_at = 0.90,      # Wait until 90%
  auto_compact = FALSE    # Disable (warnings only)
)

# Manual compaction anytime
session <- cassidy_compact(session)
```

### Memory System for Unlimited Conversations

The memory system provides persistent storage that survives compaction:

```{r, eval = FALSE}
# Save workflow state
cassidy_write_memory_file(
  "workflow_state.md",
  "# Progress\n\nCompleted Phase 3, starting Phase 4..."
)

# List memory files
files <- cassidy_list_memory_files()

# Read from memory
content <- cassidy_read_memory_file("workflow_state.md")

# Organize with subdirectories
cassidy_write_memory_file("archive/old_notes.md", "...")
cassidy_rename_memory_file("draft.md", "final.md")

# Delete when done
cassidy_delete_memory_file("temp.md")
```

**What to store in memory:**
- Workflow state and progress tracking
- Learned insights and debugging discoveries
- User preferences and project conventions
- Cross-session knowledge that should persist

**Memory + Compaction = Unlimited Conversations:**
Save important information to memory before compaction, and it persists across conversation resets.

See `vignette("managing-long-conversations")` for complete documentation.

## Roadmap

cassidyr is under active development. Current status and upcoming features:

### âœ… Complete

- **Phase 1: API Layer** - Thread creation, messaging, history retrieval
- **Phase 2: Context System** - Project, data, and file context gathering
- **Phase 3: Interactive Chat** - Shiny app with persistence, copy code button
- **Phase 4: Agentic System** - Autonomous workflows with tools and safe mode
- **Phase 5: Skills System** - Reusable methodology templates
- **Context Engineering** - Token tracking, automatic compaction, memory system, timeout handling

### â³ In Progress

- **IDE Integration**
  - `cassidy_document_function()` - Generate roxygen2 docs
  - `cassidy_explain_selection()` - Explain selected code
  - `cassidy_refactor_selection()` - Improve code quality
  - `cassidy_debug_error()` - Help debug errors

### ðŸ”® Future

- **Survey Research Tools** - EFA interpretation, methods sections, codebooks

## Design Philosophy

cassidyr is built with four core principles:

1. **Security First**: No credentials in code, secure environment variables
2. **Tidyverse Style**: Consistent naming (`cassidy_*`), pipe-friendly functions
3. **Modular Design**: Core API separate from specialized tools, easy to extend
4. **Hierarchical Memory**: Inspired by Claude Code's memory system, enabling organization-wide standards to combine with project-specific configurations

## Related Packages

- [**ellmer**](https://ellmer.tidyverse.org/): Multi-provider LLM client (OpenAI, Claude, etc.)
- [**btw**](https://posit-dev.github.io/btw/): Toolkit for connecting R and LLMs
- [**shinychat**](https://posit-dev.github.io/shinychat/): Chat UI components for Shiny

cassidyr is specifically designed for CassidyAI's unique features like persistent assistants, knowledge bases, and organizational context.

## Getting Help

- ðŸ“– [Documentation](https://jdenn0514.github.io/cassidyr/)
- ðŸ› [Report bugs](https://github.com/JDenn0514/cassidyr/issues)
- ðŸ’¬ [Ask questions](https://github.com/JDenn0514/cassidyr/discussions)

## Code of Conduct

Please note that cassidyr is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## License

MIT Â© Jacob Dennen
